---
title: "Effect of a New Inhaled Therapy on Lung Function in Mild Asthma"
author: "Natallia Halai"
toc:  true
date: today
abstract-title: " Abstract: "
abstract: "A randomized controlled trial evaluates a new inhaled corticosteroid (ICS-X) compared to a standard inhaler in patients with mild persistent asthma.
Patients are randomized to receive either ICS-X or standard therapy for 6 weeks. Lung function is assessed at baseline and at follow-up using FEV₁ (forced expiratory volume in 1 second).
The primary endpoint is improvement in FEV₁."
  
project:
  output-dir: output
  execute-dir: project

format: 
  html:
    embed-resources: true
    code-fold: true
    number-sections: true
editor: visual
editor-options:
  chunk-output-type: console
  
execute: 
  warning: false
  error: false
  freeze: true
---

# Hypothesis

-   ICS-X leads to greater improvement in FEV₁

-   Higher baseline FEV₁ is associated with less room for improvement

-   Smokers exhibit a reduced treatment response

-   A negative Age effect on improvement

```{r}
#| label: data_preparation
#| echo: false
#| output: false

library(ggplot2)
library(pastecs)
library(ggpubr); theme_set(theme_bw())
library(GGally)
library(performance)
library(see)
library(DHARMa)
library(car)
library(ggeffects)
library(MuMIn) #rsquaredLR is from here
library(knitr)
library(dplyr)
library(readr)

lung_function <- read_csv("data/processed/lung-function.csv")

str(lung_function)
# lung_function$id<-factor(lung_function$id)
lung_function$treatment_group<-factor(lung_function$treatment_group)
lung_function$smoking_status<-factor(lung_function$smoking_status)

summary(lung_function)
```

# Data Exploration

## Numerically

```{r}
#| label: tbl-exp-num
#| tbl-cap: "Correlation values for continuous variables Age and Baseline FEV"

cor_baseline<-cor(lung_function$baseline_fev1, lung_function$fev1_improvement)
cor_age<-cor(lung_function$age, lung_function$fev1_improvement)

correlation_table<-data.frame(
  Variable_1=c("Baseline FEV", "Age"),
  Variable_2=c("FEV Improvement", "FEV Improvement"),
  Corr_value=c(cor_baseline, cor_age)
)

knitr::kable(correlation_table)

```

As the first step in the data exploration correlation values were calculated for the 2 continuous variables Age and Baseline FEV as seen in @tbl-exp-num . Here we can partly confirm our hypotheses which state that Age has a negative effect on FEV improvement (corr: `r correlation_table[2, 3]` ) as well as high Baseline FEV association with less FEV improvement (corr: `r correlation_table[1,3]`).

## Graphically

```{r}
#| label: fig-exp-graph
#| fig-width: !expr 20/2.54
#| fig-height: !expr 20/2.54
#| fig-cap: ""

ggarrange(
  ggplot(lung_function, aes(x=treatment_group, y=fev1_improvement, color=treatment_group))+
    geom_point()+
    geom_boxplot(outlier.shape = NA),
  ggplot(lung_function, aes(x=baseline_fev1,y=fev1_improvement, color=treatment_group))+
    geom_point()+
    geom_smooth(method="lm"),
  ggplot(lung_function, aes(x=smoking_status, y=fev1_improvement))+
    geom_point()+
    geom_boxplot(outlier.shape=NA)
  ,
  ggplot(lung_function, aes(x=age, y=fev1_improvement, color=treatment_group))+
    geom_point()+
    geom_smooth(method="lm"), 
  common.legend = TRUE, labels=c("a)", "b)", "c)", "d)")
)

```

In @fig-exp-graph we can see different predictor variables plotted against FEV improvement response variable.