---
title: "Effect of a New Inhaled Therapy on Lung Function in Mild Asthma"
author: "Natallia Halai"
toc:  true
date: today
abstract-title: " Abstract: "
abstract: "A randomized controlled trial evaluates a new inhaled corticosteroid (ICS-X) compared to a standard inhaler in patients with mild persistent asthma.
Patients are randomized to receive either ICS-X or standard therapy for 6 weeks. Lung function is assessed at baseline and at follow-up using FEV₁ (forced expiratory volume in 1 second).
The primary endpoint is improvement in FEV₁."
  
project:
  output-dir: output
  execute-dir: project

format: 
  html:
    embed-resources: true
    code-fold: true
    number-sections: true
editor: visual
editor-options:
  chunk-output-type: console
  
execute: 
  warning: false
  error: false
  freeze: true
---

# Hypothesis

-   ICS-X leads to greater improvement in FEV₁

-   Higher baseline FEV₁ is associated with less room for improvement

-   Smokers exhibit a reduced treatment response

-   A negative Age effect on improvement

```{r}
#| label: data_preparation
#| echo: false
#| output: false

library(ggplot2)
library(pastecs)
library(ggpubr); theme_set(theme_bw())
library(GGally)
library(performance)
library(see)
library(DHARMa)
library(car)
library(ggeffects)
library(MuMIn) #rsquaredLR is from here
library(knitr)
library(dplyr)
library(readr)

lung_function <- read_csv("data/processed/lung-function.csv")

str(lung_function)
# lung_function$id<-factor(lung_function$id)
lung_function$treatment_group<-factor(lung_function$treatment_group)
lung_function$smoking_status<-factor(lung_function$smoking_status)

summary(lung_function)
```

# Data Exploration

## Numerically

```{r}
#| label: tbl-exp-num
#| tbl-cap: "Correlation values for continuous variables Age and Baseline FEV"

cor_baseline<-cor(lung_function$baseline_fev1, lung_function$fev1_improvement)
cor_age<-cor(lung_function$age, lung_function$fev1_improvement)

correlation_table<-data.frame(
  Variable_1=c("Baseline FEV", "Age"),
  Variable_2=c("FEV Improvement", "FEV Improvement"),
  Corr_value=c(cor_baseline, cor_age)
)

knitr::kable(correlation_table)

```

As the first step in the data exploration correlation values were calculated for the 2 continuous variables Age and Baseline FEV as seen in @tbl-exp-num . Here we can partly confirm our hypotheses which state that Age has a negative effect on FEV improvement (corr: `r correlation_table[2, 3]` ) as well as high Baseline FEV association with less FEV improvement (corr: `r correlation_table[1,3]`).

## Graphically

```{r}
#| label: fig-exp-graph
#| fig-width: !expr 20/2.54
#| fig-height: !expr 20/2.54
#| fig-cap: ""

ggarrange(
  ggplot(lung_function, aes(x=treatment_group, y=fev1_improvement, color=treatment_group))+
    geom_point()+
    geom_boxplot(outlier.shape = NA),
  ggplot(lung_function, aes(x=baseline_fev1,y=fev1_improvement, color=treatment_group))+
    geom_point()+
    geom_smooth(method="lm"),
  ggplot(lung_function, aes(x=smoking_status, y=fev1_improvement))+
    geom_point()+
    geom_boxplot(outlier.shape=NA)
  ,
  ggplot(lung_function, aes(x=age, y=fev1_improvement, color=treatment_group))+
    geom_point()+
    geom_smooth(method="lm"), 
  common.legend = TRUE, labels=c("a)", "b)", "c)", "d)")
)

```

In @fig-exp-graph we can see different predictor variables plotted against FEV improvement response variable.

# Model Fitting

## Create Candidate Models

```{r}
#| label: tbl-cand-models
#| tbl-cap: "Model performance comparison"
#| echo: false

model.00<-lm(fev1_improvement~1, lung_function)
model.01<-lm(fev1_improvement~baseline_fev1, lung_function)
model.02<-lm(fev1_improvement~baseline_fev1+treatment_group, lung_function)
model.03<-lm(fev1_improvement~baseline_fev1+treatment_group+age, lung_function)
model.04<-lm(fev1_improvement~baseline_fev1+treatment_group+age+smoking_status, lung_function)

knitr::kable(compare_performance(model.00, model.01, model.02, model.03, model.04), digits=3)

```

We will be choosing the model.04 which has all the predictor variables included as it explains the most variance in the data.

## Check chosen models diagnostics

```{r}
#| label: fig-diagnostic-model
#| fig-cap: "Diagnostic plots"
#| fig-width: !expr 25/2.54
#| fig-height: !expr 25/2.54

check_model(model.04)
```

In @fig-diagnostic-model we can see that there is no high collinearity between the variables, the homoscedasticity assumption is also met as well as the Normality (normal residuals distribution) and Linearity(trends in the residuals). Also we can detect no outliers/influencial points in the Cook's plot.

```{r}

```
